Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-11-23 doc/manuscript.tex
discussion drafted

\newcommand{\aprcomment}[1]{{\textcolor{blue}{Comment: #1}}}
\newcommand{\E}{\mathbb{E}}

@@

\begin{abstract}
    .
\end{abstract}

@@

Many phenotypic traits are under stabilizing selection
\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This motivates
stabilizing selection around a shared optimum as an appropriate null models for
the dynamics of allele frequencies affecting polygenic traits
\citep{yair2022population}. Stabilizing selection results in distinct dynamics
for trait-affecting alleles, characterized by symmetric underdominance
\citep{robertson1956effect,keightley1988quantitative}, that is, selection
against the minor allele. Because admixture and introgression events are common
in natural populations, it will be important to understand the effects of gene
flow on the architecture of complex traits.  \aprcomment{or\dots``studies of
complex traits in admixed populations rely on this understanding, or
something''}

After admixture between diverged populations, the additive genetic variance of
a trait rapidly increases. The genetic variance after admixture depends on both
the existing genetic variances within the parental population and their
divergence at trait-affecting loci, measured by $F_2$
(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to fairly quickly
decay to background levels. During this process, introgressed and pre-existing
trait-affecting alleles are replaced by new mutations, turning over the genetic
architecture of the trait (Figure\ref{fig:toy-admixture}). Selection occurs
against the minor alleles at trait-affecting loci, so that introgressed alleles
from the minor parental ancestry, whether derived or reintroduced ancestral
alleles, are selected against (Figure~\ref{fig:linkage-pred}).

@@

The expected pattern of shared ancestry deserts under stabilizing selection
differs from models of both deleterious load and incompatibilities. Load-based
models predict that haplotypes that have accumulated more deterious mutations,
e.g., from a population with small long-term effective population size, will be
selected against under either direction of gene flow. Introgressed ancestry at
a given selected locus will decrease in frequency in one introgression scenario
and increase the other. This may explain the replacement of MT and Y chromosome
DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
introgression \citep{posth2017deeply,petr2020evolutionary}.

\citet{muller1942isolating} originally hypothesized that hybrid
incompatibilities should not often form within a single locus, but instead the
two or more loci involved will most often be found at distant or unlinked loci
in genome. Theory predicts that hybrid incompatibilities are resolved via
selection against the minor parental ancestry [cite]. Depending on the
direction of introgression, different incompatibilities alleles will be
selected against and ancestry deserts will form around those different loci.
While such a model predicts selection against introgressed ancestry, similar to
the stabilizing selection model, an important distinction is that deserts are
not expected to overlap with bidirectional gene flow.

\aprcomment{any data about distributions of incompatibility loci in the
genome?}

\aprcomment{Second-to-last paragraph}
Turning to observations from human and Neanderthal genomes
\citep{harris2023diverse}

\aprcomment{last paragraph}
\begin{itemize}
    \item Selection in structured populations with ongoing migration or
        meta-populations -- SFS approach can handle this -- prospects for the
        future
\end{itemize}
\aprcomment{end on a comment about the usefulness of the SFS approach and the
promise for future studies of dynamics of the architecture of complex traits
with non-trivial demographic scenarios -- possible that this should be earlier,
and we can include caveats there. Maybe second paragraph of discussion.}

\aprcomment{don't end on this, or might not need to include it -- caveats can
come earlier?}
\begin{itemize}
    \item Some conjectures about hybrid zones - single-locus ``incompatibilities''
    \item Pleiotropy\dots!!!
\end{itemize}

>>>>

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\E}{\mathbb{E}}

@@

\begin{abstract}

\end{abstract}

@@

Many phenotypic traits are under stabilizing selection
\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This has
motivated using stabilizing selection around a shared optimum as a null model
for the dynamics of alleles affecting polygenic traits
\citep{yair2022population}. Stabilizing selection has a distinct effect on
trait-affecting alleles, characterized by symmetric underdominance
\citep{robertson1956effect,keightley1988quantitative}, that is, selection
against the minor allele. There has been some previous work on how migration
shapes the genetic variance of a trait under stabilizing selection
\aprcomment{[cite]}, but most studies consider single populations, often at
steady state. Admixture and introgression events are common in natural
populations, so understanding their effects on the architecture of complex
traits is needed.

Here, we have shown that multi-population, non-equilibrium approaches for the
site frequency spectrum with underdominance can be used to model the additive
genetic variance of a trait under stabilizing selection. While this approach
ignores biologically relevant aspects such as linkage between sites and
pleiotropy, it can still provide important insights into the dynamics of trait
architectures. By modeling allelic contributions to the trait as additive, it
allows for the decomposition of contributions from alleles of different
origins, either by source population or mutation time. It may therefore be
useful for understanding the expected contributions of introgressed variants to
complex traits, for example from Neanderthal-introgressed mutations in humans
\citep{wei2023lingering}. In the limited scenarios explored here, we see that
the expected contributions of introgressed variants to heritability are
complicated even in the purely additive case, depending on the distribution of
effect sizes, demographic history, and the time since admixture
(Figures~\ref{fig:neand-to-human}, \ref{fig:human-to-neand}).

\aprcomment{While we have focused on the relatively simple scenario of two
diverged populations recently exchanging migrants, SFS methods presented here
can incorporate continuous migration, more than two populations, or changes in
the trait optimum. We leave these for future work\dots}

After admixture between diverged populations, the additive genetic variance of
a trait rapidly increases. The genetic variance after admixture depends on both
the existing genetic variances within the parental population and their
divergence at trait-affecting loci, measured by $F_2$
(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to decay fairly
quickly to background levels. During this process, introgressed and
pre-existing trait-affecting alleles are replaced by new mutations, turning
over the genetic architecture of the trait (Figure\ref{fig:toy-admixture}).
Selection occurs against the minor alleles at trait-affecting loci, so that
introgressed alleles from the minor parental ancestry, whether derived or
reintroduced ancestral alleles, are selected against
(Figure~\ref{fig:linkage-pred}).

@@

The expected pattern of shared ancestry deserts under stabilizing selection
differs from models of both deleterious load and incompatibilities. Load-based
models predict that haplotypes that have accumulated more deterious mutations,
e.g., from a population with small long-term effective population size, will be
selected against under either direction of gene flow. Introgressed ancestry at
a given selected locus will decrease in frequency in one introgression scenario
and increase the other. This may explain the replacement of MT and Y chromosome
DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
introgression \citep{posth2017deeply,petr2020evolutionary}.
%However, this is
%not the general pattern observed when comparing introgressed ancestry in the
%autosomal genomes of humans and Neanderthals \citep{harris2023diverse}.

The classical model of Bateson-Dobzhansky-Muller incompatibilities (BDMIs)
\citep{bateson1909heredity,dobzhansky1936studies,muller1942isolating} explains
the accumulation of reproductive isolation through negative epistatic
interactions that are exposed in hybrids. \citet{muller1942isolating}
hypothesized that such BDMIs should most often form between distant or unlinked
loci, instead of within single or tightly linked loci. Because theory predicts
that hybrid incompatibilities are resolved via selection against the minor
parental ancestry \citep{moran2021genomic}, ancestry deserts should form in
different genomic regions as the different incompatibility alleles are selected
against, depending on the direction of introgression. While such a model
predicts selection against introgressed ancestry, similar to that of
stabilizing selection, an important distinction is that deserts due to
selection against incompatibility loci are not expected to overlap under
bidirectional gene flow. While there is little empirical data on the
distribution of BDMIs, studies point to interacting BDMI alleles being unlinked
\citep[][e.g.,]{presgraves2003fine} and an asymmetry in the alleles under
selection in different introgression scenarios
\citep{maheshwari2011genetics,moran2021genomic}.

Finally, turning to the distribution of introgressed ancestry deserts in humans
and Neanderthals, \citet{harris2023diverse} find that regions of depleted
Neanderthal ancestry in humans overlap more than would be expected by chance
with regions lacking human-introgressed alleles in the Altai Neanderthal.
Human-introgressed ancestry in the Neanderthal genome is also depleted in
functional regions, as observed in humans \citep{sankararaman2014genomic}.
\citet{harris2023diverse} propose that epistatic interactions between
introgressed alleles and the recipient backgrounds could drive these patterns,
which they interpret as evidence for the inital process of speciation between
humans and Neanderthals. However, this observation of shared ancestry deserts
does not match expectations under a classic model of BDMIs, as argued above.
Instead, at least some of the pattern may be due to stabilizing selection
acting on complex traits, such as gene regulation.  Importantly, such
overlapping ancestry deserts are expected even when a trait is under
stabilizing selection for the same phenotypic optimal value. The selective
causes of ancestry deserts in both humans and Neanderthals remain unknown.
Stabilizing selection will prove a useful null model for testing scenarios of
epistasis, incompatibilities and adaptive introgression.
